Servis: A dependently typed DSL for web APIs

### 2016-08-22 paper/paper.tex
work

\subsection{Overlapping routes}
Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{needed}.
It would be nice if we could verify at compile time that no routes in an API specification can overlap.

@@

\subsection{The Universe pattern}
The Servis DSL makes use of the Universe pattern to map an API description to types of handlers.
 A universe is a type \mintinline{idris}{U : Type} which contains names for types and a type family \mintinline{idris}{el : U -> Type} that assigns to every name \mintinline{idris}{a : U} the type of its elements \mintinline{idris}{el a : Type} \cite{generic}\@.

@@

-- we define a value of API
api : API
api = Const "users" :> Capture Int "userId" :> Outputs (Get USER)

@@

--  el api evaluates to  Int -> IO User
handler : el api
handler = getUser
 \end{minted}
\subsection{The DSL}
So what does this type \mintinline{idris}{API} look like?
\subsubsection{Handler}
Before we will look at routes, lets think of what an endpoint should look like.
We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
specify a request body type, and a return type.

@@

The universe implementation of an API is the type of a list of handlers.
Every handler is of a different type though, so instead of using an ordinary list,
we use an heterogenous list, where each element \mintinline{idris}{x} in \mintinline{idris}{xs}
is of type \mintinline{idris}{el x}.
\begin{minted}
( Universe capture
, Universe query
, Universe req
, Universe res
) => Universe (API capture query req res) where
  el (OneOf xs) = HVect (map el xs)

\subsection{A simple interpreter: Documentation generation}
Here we will show a simple interpretation of the DSL, by generating some basic documentation.
It's basically a function \mintinline{idris}{docs : API -> String}. 
\subsection{An interpreter for an HTTP Server}\label{sec:server}
A more advanced example where we define two interpreters. One 'interpreter' that generates types from our DSL such that we can type-check our server handler implementations. (The universe pattern), and another interpreter that routes requests to these handler implementations.
\subsubsection{Checking explicit status codes}
Shows that we can enforce explicit status codes

@@

||| A Proof that the GET dtor is injective over its requestType
injGET : (GET x = GET y) -> x = y
injGET Refl = Refl

||| A proof that the POST dtor is injective over its requestType
injPOSTReq : (POST requestType1 responseType1 = POST requestType2 responseType2) -> requestType1 = requestType2
injPOSTReq Refl = Refl

||| A Proof that the POST dtor is injective over its responseType
injPOSTResp : (POST requestType1 responseType1 = POST requestType2 responseType2) -> responseType1 = responseType2
injPOSTResp Refl = Refl

||| A Proof that a GET is not a POST
getNotPost : (GET responseType = POST requestType responseType1) -> Void
getNotPost Refl impossible

||| A proof that GET is congruent
cong_GET : (prf : responseType1 = responseType2) -> GET responseType1 = GET responseType2
cong_GET = cong

||| A proof POST is congruent
cong_POST : (prf2 : responseType1 = responseType2) -> (prf1 : requestType1 = requestType2) -> POST requestType1 responseType1 = POST requestType2 responseType2
cong_POST Refl Refl = Refl

(DecEq req, DecEq res) => DecEq (Handler req res) where
  decEq (GET responseType1) (GET responseType2) =
    (case decEq responseType1 responseType2 of
          (Yes prf) => Yes (cong prf)
          (No contra) => No ((\h => contra (injGET h))))
  decEq (POST requestType1 responseType1) (POST requestType2 responseType2) =
    (case decEq requestType1 requestType2 of
          (Yes prf1) => (case decEq responseType1 responseType2 of
                             (Yes prf2) => Yes (cong_POST prf2 prf1)
                             (No contra) => No (\h => contra (injPOSTResp h)))
          (No contra) => No (\h => contra (injPOSTReq h)))
  decEq (GET _) (POST _ _) = No getNotPost
  decEq (POST _ _) (GET _) = No (negEqSym getNotPost)
data DisjointPP : PathPart capture query -> PathPart capture query -> Type where
  ConstD : Not (str = str') -> DisjointPP (Const str) (Const str')

data DisjointPath : Path capture query req res -> Path capture query req res -> Type where
  OutputsD : Not (handler1 = handler2) -> DisjointPath (Outputs handler1) (Outputs handler2)
  PathBase : DisjointPP pp1 pp2 -> DisjointPath (pp1 :> p1) (pp2 :> p2)
  PathStep : DisjointPath p1 p2 -> DisjointPath (p :> p1) (p :> p2)

data DisjointAPI : List (Path capture query req res) -> Type where
  Base : DisjointAPI []
  Step : (x : Path capture query req res) -> All (DisjointPath x) xs -> DisjointAPI (x::xs)
\end{minted}
\subsection{Pi-types}
Adds dependent pairs to our DSL such that we can describe dependencies between query params, captures and return types.  Example:

@@

\end{document}

>>>>

\subsection{Overlapping routes}
Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
It would be nice if we could verify at compile time that no routes in an API specification can overlap.

@@

\subsection{The Universe pattern}
The Servis DSL makes use of the Universe pattern to map an API description to types of handlers.
 A universe is a type \mintinline{idris}{U : Type} which contains names for types and a type family \mintinline{idris}{el : U -> Type} that assigns to every name \mintinline{idris}{a : U} the type of its elements \mintinline{idris}{el a : Type} \cite{Altenkirch}\@.

@@

-- we define a value of API
api : API
api = Const "users" :> Capture Int "userId" :> Outputs (GET User)

@@

--  el api evaluates to  Int -> IO User
handler : el api
handler = getUser
 \end{minted}
\subsection{The DSL}
So what does this type \mintinline{idris}{API} look like?
\subsubsection{Handler}\label{sec:handler}
Before we will look at routes, lets think of what an endpoint should look like.
We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
specify a request body type, and a return type.

@@

The universe implementation of an \mintinline{idris}{API} is the type of a list of handlers.
Every handler is of a different type though, so instead of using an ordinary list,
we use an heterogenous list, where each element \mintinline{idris}{x} in \mintinline{idris}{xs}
is of type \mintinline{idris}{el x}.
\begin{minted}
( Universe capture
, Universe query
, Universe req
, Universe res
) => Universe (API capture query req res) where
  el (OneOf xs) = HVect (map el xs)
\end{minted}.

\subsection{An example API}
Using this DSL, we will now describe an API for interacting with users in a database.
We will start of with defining our domain objects, and intercations we could do with them.

\begin{listing}
\begin{minted}{idris}
record User where
  constructor MkUser
  id : Int
  name : String
  email : String
  password : String

findUserByName : String -> IO (List User)
getUsers : (limit : Int) -> IO (List User)
getUserById : (id : Int) -> IO User
addUser : User -> IO ()
\end{minted}
\label{domain}
\caption{The User object and its interactions}
\end{listing}

Next, we will describe the HTTP API for interacting with these users.  This includes
the need of coming up with universes for captures, query params, request body, and response.

\begin{listing}
\begin{minted}{idris}
data QueryUniverse
  = QSTRING
  | QINT
Universe QueryUniverse where
  el QSTRING = String
  el QINT = Int

data CaptureUniverse
  = CINT
Universe CaptureUniverse where
  el CINT = Int

-- Note: We use the same universe for requests and responses.
data ReqResUniverse
  = USER
  | LIST ReqResUniverse
Universe ReqResUniverse where
  el USER = User
  el (LIST a) = List (el a)

userAPI : API QueryUniverse CaptureUniverse ReqResUniverse ReqResUniverse
userAPI = OneOf
  [ Const "users" :> Capture "user_id" CINT :> Outputs (GET USER)
  , Const "users" :>
    Const "find" :>
    QueryParam "email" QSTRING :>
    Outputs (GET (LIST USER))
  , Const "users" :> Outputs (POST USER)
  , Const "users" :>
    QueryParam "limit" QINT :>
    Outputs (GET (LIST USER))
  ]
\end{minted}
\label{code:api}
\caption{API definition for users}
\end{listing}

\subsection{A simple interpreter: Documentation generation}
Writing an interpreter for generating documentation is very easy. Our API description isn't some type-level definition, it's just a plain old datatype that can be passed to functions and can be manipulated.

We start by introducing an interface for docs:
\begin{minted}{idris}
interface HasDocs a where
  docs :: a -> String
\end{minted}
Now, we just define an implementation for \mintinline{idris}{API} for this interface and we're done. Which is just calling \mintinline{idris}{docs} recursively on all its children.

\begin{minted}{idris}
(HasDocs req, HasDocs res) => HasDocs (Handler req res) where
  docs (GET responseType) =
    "## GET\nResponse type: " ++ docs responseType

  docs (POST requestType responseType) =
    "## POST\nRequest type: "  ++ docs requestType ++ "\n" ++
    "Response type: " ++ docs responseType ++ "\n"

renderParams : HasDocs query => List (String, query) -> String
renderParams = renderParams' . reverse
  where
    renderParams' : List (String, query) -> String
    renderParams' [] = ""
    renderParams' ((a, b) :: []) =
      "?" ++ a ++ "=<"++ docs b ++ ">"
    renderParams' ((a, b) :: xs) =
      renderParams' xs ++ "&" ++ a ++ "=<"++ docs b ++ ">"

renderPath : 
  ( HasDocs capture
  , HasDocs query
  , HasDocs req
  , HasDocs resp) => Path capture query req resp -> String -> String
renderPath (Outputs x) params = 
    params ++ "\n" ++ docs x
renderPath ((Const x) :> y) params =
    "/" ++ x ++ renderPath y params
renderPath ((Capture name type) :> y) params =
  "/<" ++ name ++ ":" ++ docs type ++ ">" ++ renderPath y params
renderPath _ _ = ""

( HasDocs capture
, HasDocs query
, HasDocs req
, HasDocs resp
) => HasDocs (Path capture query req resp) where
  docs path =
    "# " ++ renderPath path (renderParams . getParams $ path) ++ "\n"
    where

      getParams : Path capture query req resp -> List (String, query)
      getParams (Outputs x) = []
      getParams ((Const x) :> y) = getParams y
      getParams ((Capture name type) :> y) = getParams y
      getParams ((QueryParam name type) :> y) = (name, type) :: getParams y

( HasDocs capture
, HasDocs query
, HasDocs req
, HasDocs resp
) => HasDocs (API capture query req resp) where
  docs (OneOf (x::[])) = docs x
  docs (OneOf (x::y::xs)) = docs x ++ "\n" ++ oneOf (y::xs)
\end{minted}

Now \mintinline{idris}{docs userAPI}, assuming that the universes implement
the \mintinline{idris}{HasDocs} interface, will evaluate to a very basic form of documentation:
\begin{verbatim}
# /users/<user_id:CINT>
## GET
 Response type: USER
# /users/find?email=<QSTRING>
## GET
 Response type: LIST USER

# /users?limit=<QINT>
## GET
 Response type: LIST USER

# /users
## POST
Request type: USER
Response type: USER

# /users?limit=<QINT>
## GET
 Response type LIST USER
\end{verbatim}

\subsection{An interpreter for an HTTP Server}\label{sec:server}
Now that we've shown that it is easy to manipulate APIs because they're just datatypes, it's time for a more complicated example. Though it is cool that we can generate documentation from an API description, the useful part that we want to implement is an HTTP Server.

\subsubsection{A type for web applications}
Lets first ask the question how a web server should look like in Idris. In the Servant paper, they describe the type of http handlers in the \texttt{wai} library, which looks like \cite{servant}:
\begin{minted}{haskell}
type Application = Request -> IO Response
\end{minted}.

We will more-or-less use this representation to model our HTTP handlers. they are  simply a function from request to response.  The actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.

Instead of having two abstract types \mintinline{idris}{Request} and \mintinline{idris}{Response}, we will assume for simplicity that a \mintinline{idris}{Response} is just a \mintinline{idris}{String}, and that a \mintinline{idris}{Request} is an \mintinline{idris}{URL}, and a requestBody of type \mintinline{idris}{Maybe String}.

\begin{minted}{idris}
Application : Type
Application = (url : URL) -> (requestBody : Maybe String) -> IO String
\end{minted}

Just like in the Servant paper, we will need to modify this type a little bit more to suit our needs. We need to be able to express that an \mintinline{idris}{Application} does not match in case of a routing mismatch, such that we can try another route encompassed in \mintinline{idris}{OneOf}.

Instead of returning  a value of \mintinline{idris}{IO String}, we can return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.

\begin{minted}{idris}
RoutingApplication : Type
RoutingApplication = (url : URL) ->
                     (requestBody : Maybe String)
                     -> Maybe (IO String)
\end{minted}

\subsubsection{Decoding and encoding data}
In section~\ref{sec:handler}, we showed that if we want to implement some decoder, we would have to resort
to the universe pattern. So that is what we will do exactly here.

As an example, we will show how we would decode a queryparam. All other decoders follow the same pattern.
What we want to do is, given a \mintinline{idris}{queryParam : String}, and a value \mintinline{idris}{v : u},
where \mintinline{idris}{u} is the universe of query param types, and  \mintinline{idris}{v} a name of a type in that universe,
perform the correct decoding to get a value of type \mintinline{idris}{el v}.

We can summarize this behaviour in an interface:
\begin{minted}{idris}
interface Universe u => FromQueryParam u where
  fromQueryParam : (v : u) -> String -> Maybe (el v)
\end{minted}

similarly, we have interfaces for captures and request bodies:
\begin{minted}{idris}
interface Universe u => FromCapture u where
  fromCapture : (v : u) -> String -> Maybe (el v)

interface Universe u => FromRequest u where
  fromRequest : (v : u) -> String -> Maybe (el v)
\end{minted}

An implementation of \mintinline{idris}{FromQueryParam} for the universe \mintinline{QueryUniverse} in Listing~\ref{code:api}, would look like this:

\begin{minted}{idris}
FromQueryParam QueryU where
  fromQueryParam QINT str = maybeParseInt str
\end{minted}

The fact that we can pattern match on the universe lets us inspect what type we need to return,
and select the right decoder accordingly.

We can use the same technique for encoding responses:
\begin{minted}{idris}
interface Universe u => ToResponse u where
  toResponse : (v : u) -> el v -> String

\end{minted}

For example, we could use this to encode values in the \mintinline{idris}{ReqResUniverse} universe from
Listing~\ref{code:api}:
\begin{minted}{idris}
  ToResponse ReqResUniverse where
    toResponse USER (MkUser id name email password) =
      "MkUser " ++ show id ++ " " ++ show name
    toResponse (LIST elem) xs =
      foldr (++) ""  . map (toResponse elem) $ xs
\end{minted}

\subsubsection{Routing requests}
We will define the following interface to implement routing requests:
\begin{minted}{haskell}
interface Universe u => HasServer u where
  route : (v : u) -> (handler : el v) -> RoutingApplication
\end{minted}
Or if we expaned the definition \mintinline{idris}{RoutingApplication}:
\begin{minted}{haskell}
interface Universe u => HasServer u where
  route : (v : u) ->
          (handler : el v) ->
          (url : URL) ->
          (requestBody : Maybe String)
          -> Maybe (IO String)
\end{minted}

The idea of the router is, that given some part of the API \mintinline{idris}{v : u}, and a handler of type \mintinline{idris}{el v}, build an application that can either route succesfully or fail. Preferably, the application uses the handler parameter in its implementation of course.

Lets start by defining an \mintinline{idris}{HasServer} instance for \mintinline{idris}{Handler}. Its task is to simply encode responses to Strings, and to handle request bodies to calculate responses
by calling the handler and encoding its result.
\begin{minted}{idris}
(FromRequest req, ToResponse res) => HasServer (Handler req res) where
  -- route : Handler req res -> 
  --         (handler : IO res) ->
  --         URL -> Maybe String -> Maybe (IO String)
  ||| simply turn an (IO res) into an (IO String) 
  ||| using toResponse
  route (GET responseType) handler url requestBody =
    Just (map (toResponse responseType) handler)
  route (POST requestType responseType) handler url requestBody = do
    -- see if there is a requestBody. otherwise fail
    body <- requestBody
    -- try to decode the request. this might fail and return Nothing
    request <- fromRequest requestType body
    -- apply the decoded request to the handler, and encode the result
    Just (map (toResponse responseType) (handler request))
\end{minted}

@@

\end{minted}
\subsection{Pi-types}
Adds dependent pairs to our DSL such that we can describe dependencies between query params, captures and return types.  Example:

@@

\bibliography{sources}
\bibliographystyle{plain}
\end{document}
